Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2018-06-27 workflow/workflow.Rmd
Change code to work with latest package versions

---
title: "Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference"
author: 
  - name: Fanny Perraudeau
    affiliation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
  - name: Davide Risso
    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine, New York, NY
  - name: Kelly Street
    affilitation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
  - name: Elizabeth Purdom
    affiliation: Department of Statistics, University of California, Berkeley, Berkeley, CA
  - name: Sandrine Dudoit
    affiliation: Division of Biostatistics and Department of Statistics, University of California, Berkeley, Berkeley, CA

abstract: Novel single-cell transcriptome sequencing assays allow researchers to measure gene expression levels at the resolution of single cells and offer the unprecendented opportunity to investigate at the molecular level fundamental biological questions such as stem cell differentiation or the discovery and characterization of rare cell types. However, such assays raise challenging statistical and computational questions and require the development of novel methodology and software. Using stem cell differentiation in the mouse olfactory epithelium as a case study, this integrated workflow provides a step-by-step tutorial to the methodology and associated software for the following four main tasks:(1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and pseudotimes; and (4) differential expression analysis along lineages.

keywords: single-cell, RNA-seq, normalization, dimensionality reduction, clustering, lineage inference, differential expression, workflow

bibliography: ref.bib

vignette: >
    %\VignetteIndexEntry{A worfklow for low-level analyses of scRNA-seq data}
    %\VignetteEngine{knitr::rmarkdown}
output: BiocWorkflowTools::f1000_article
---

@@

library(BiocStyle)
library(knitr)
knitr::opts_chunk$set(cache=TRUE, error=FALSE, message=FALSE, warning=FALSE)
knitr::opts_chunk$set(fig.align="center", fig.width=5, fig.height=5)
```

@@

```{r packages}
# GitHub
library(slingshot)

# Bioconductor
library(BiocParallel)
library(clusterExperiment)
library(scone)
library(zinbwave)
library(SingleCellExperiment)

@@

Note that in order to successfully run the workflow, we need the devel versions
of the Bioconductor packages `scone (>=1.1.2)`, `zinbwave (>=0.99.6)`, and `clusterExperiment (>=1.6)`. We recommend running Bioconductor 3.6 (currently the devel version; see https://www.bioconductor.org/developers/how-to/useDevel/).

@@

@@

Throughout the workflow, we use the class `SummarizedExperiment` to keep track of the counts and their associated metadata within a single object. The cell-level metadata contain quality control measures, sequencing batch ID, and cluster and lineage labels from the original publication [@Fletcher2017]. Cells with a cluster label of `-2` were not assigned to any cluster in the original publication.

@@

dataObj <- SummarizedExperiment(assays = list(counts = E),
                           colData = metadata)
dataObj
```

@@

mfilt <- metric_sample_filter(assay(dataObj), 
                              nreads = colData(dataObj)$NREADS,
                              ralign = colData(dataObj)$RALIGN,
                              pos_controls = rownames(dataObj) %in% hk,
                              zcut = 3, mixture = FALSE,
                              plot = TRUE)
```

@@

```{r}
# # Filtering to top 1,000 most variable genes
vars <- rowVars(log1p(assay(dataObj)))
names(vars) <- rownames(dataObj)
vars <- sort(vars, decreasing = TRUE)
# dataObj <- dataObj[names(vars)[1:1000],]

#alternative script:
dataObj<-makeFilterStats(dataObj,filterStats="var")
dataObj<-filterData(dataObj,percentile=1000,filterStats="var")

@@

@@

@@

```{r addW}
#reducedDim(dataObj,"zinbwave50")<-W
```

@@

@@

@@

@@

@@

@@

Cluster name | Description | Color | Correspondence
-------------|-------------|-------| ----------
c1 | HBC | blue | original 1, 5
c2 | GBC | green | original 2, 3, 11
c3 | mSUS | red | original 4, 7
c4 | Contaminants | orange | original -2 
c5 | mOSN | purple | original 9, 12
c6 | Immature Neuron | brown | original 10, 14
c7 | MV | cyan | original 15

@@

@@

@@

```{r}
pseudoCe <- dataObj[,!primaryClusterNamed(dataObj) %in% c("-1", "c1")]
X <- reducedDim(pseudoCe,type="zinbwave")
mds <- cmdscale(dist(X), eig = TRUE, k = 4)
X <- mds$points
lineages <- getLineages(X, clusterLabels = primaryClusterNamed(pseudoCe), start.clus = "c1")
```
EP: See my email. I don't get this subsetting error on my machine or on TravisCI. I do get on SCF Servers which are running R3.4.1 and the older version of bioconductor. Can you check this on your machine?

@@

@@

```{r tree, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space, with connecting lines between cluster centers representing the inferred global lineage structure."}
colorCl<-convertClusterLegend(pseudoCe,whichCluster="primary",output="matrixColors")[,1]
pairs(lineages, type="lineages", col = colorCl)
```

@@

```{r fitgam}
t <- pseudotime(lineages)[,1]
y <- transformData(pseudoCe)
gam.pval <- apply(y,1,function(z){
  d <- data.frame(z=z, t=t)
  tmp <- gam(z ~ lo(t), data=d)
  p <- summary(tmp)[4][[1]][1,5]
  p
})
```

@@

```{r heatmapsignificant, fig.cap="DE: Heatmap of the normalized expression measures for the 100 most significantly DE genes for the neuronal lineage, where rows correspond to genes and columns to cells ordered by pseudotime."}
topgenes <- names(sort(gam.pval, decreasing = FALSE))[1:100]
orderSamples(pseudoCe)<-order(t,na.last=TRUE)
plotHeatmap(pseudoCe[topgenes,!is.na(t)], clusterSamplesData = "orderSamplesValue", breaks = .99)
```

>>>>

---
title: "Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference"
author:

  - name: Fanny Perraudeau
    affiliation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
  - name: Davide Risso
    affiliation: Division of Biostatistics and Epidemiology, Department of Healthcare Policy and Research, Weill Cornell Medicine, New York, NY
  - name: Kelly Street
    affilitation: Graduate Group in Biostatistics, University of California, Berkeley, Berkeley, CA
  - name: Elizabeth Purdom
    affiliation: Department of Statistics, University of California, Berkeley, Berkeley, CA
  - name: Sandrine Dudoit
    affiliation: Division of Biostatistics and Department of Statistics, University of California, Berkeley, Berkeley, CA
abstract: Novel single-cell transcriptome sequencing assays allow researchers to measure gene expression levels at the resolution of single cells and offer the unprecendented opportunity to investigate at the molecular level fundamental biological questions such as stem cell differentiation or the discovery and characterization of rare cell types. However, such assays raise challenging statistical and computational questions and require the development of novel methodology and software. Using stem cell differentiation in the mouse olfactory epithelium as a case study, this integrated workflow provides a step-by-step tutorial to the methodology and associated software for the following four main tasks:(1) dimensionality reduction accounting for zero inflation and over-dispersion and adjusting for gene and cell-level covariates; (2) cell clustering using resampling-based sequential ensemble clustering; (3) inference of cell lineages and pseudotimes; and (4) differential expression analysis along lineages. 
keywords: single-cell, RNA-seq, normalization, dimensionality reduction, clustering, lineage inference, differential expression, workflow
bibliography: ref.bib
vignette: >
    %\VignetteIndexEntry{A worfklow for low-level analyses of scRNA-seq data}
    %\VignetteEngine{knitr::rmarkdown}
output: BiocWorkflowTools::f1000_article
editor_options: 
  chunk_output_type: console
---

@@

library(BiocStyle)
library(knitr)
knitr::opts_chunk$set(cache=FALSE, error=FALSE, message=FALSE, warning=FALSE)
knitr::opts_chunk$set(fig.align="center", fig.width=5, fig.height=5)
```

@@

```{r packages}
# Bioconductor
library(BiocParallel)
library(clusterExperiment)
library(scone)
library(zinbwave)
library(SingleCellExperiment)
library(slingshot)

@@

Note that in order to successfully run the workflow, we need the devel versions
of the Bioconductor packages `scone (>=1.5.0)`, `zinbwave (>=1.3.2)`, and `clusterExperiment (>=2.1.3)`. We recommend running Bioconductor 3.8 (currently the devel version; see https://www.bioconductor.org/developers/how-to/useDevel/).

@@

DR: Done, but let's double-check again before the final submission.

@@

Throughout the workflow, we use the class `SingleCellExperiment` to keep track of the counts and their associated metadata within a single object. The cell-level metadata contain quality control measures, sequencing batch ID, and cluster and lineage labels from the original publication [@Fletcher2017]. Cells with a cluster label of `-2` were not assigned to any cluster in the original publication.

@@

dataObj <- SingleCellExperiment(assays = list(counts = E),
                           colData = metadata)
dataObj
```

@@

mfilt <- metric_sample_filter(counts(dataObj), 
                              nreads = colData(dataObj)$NREADS,
                              ralign = colData(dataObj)$RALIGN,
                              pos_controls = rownames(dataObj) %in% hk,
                              zcut = 3, mixture = FALSE,
                              plot = TRUE)
```

@@

```{r}
# # Filtering to top 1,000 most variable genes
vars <- rowVars(log1p(counts(dataObj)))
names(vars) <- rownames(dataObj)
vars <- sort(vars, decreasing = TRUE)
# dataObj <- dataObj[names(vars)[1:1000],]

#alternative script:
dataObj<-makeFilterStats(dataObj,filterStats="var", transFun = log1p)
dataObj<-filterData(dataObj,percentile=1000,filterStats="var")

@@

DR: Fixed, the new code was computing the variance in linear space instead of using the log.

DR: Remember to remove the old code and the length() and setdiff() from above.

@@

DR: Agree, I'm not even sure normalized values are that useful for visualization... a simple scaling by total counts would probably achieve the same... but since we have them...

@@

@@

DR: A lot of these comments are now outdated, as the new version of clusterExperiment fully support multiple assays. Similarly, the weights are now a reality and also supported by clusterExperiment. We should probably add a section about it?

@@

DR: I fixed the most variable genes. Is there a way to check the older results? Where do I find the saved objects?

@@

DR: it's pretty similar, but we still get 7 clusters that look more or less the same as before.

@@

DR: Now it looks pretty much the same.

@@

DR: It is possible that zinbwave does not return the same exact W. There is no guarantee of a global maximum, but only local. The plot looks more or less the same, but just flipped.

@@

Cluster name | Description | Color | Correspondence
-------------|-------------|-------| ----------
c1 | HBC | blue | original 1, 5
c2 | GBC | green | original 2, 3, 11
c3 | mSUS | red | original 4, 7
c4 | Contaminants | orange | original -2 
c5 | mOSN | purple | original 9, 12
c6 | Immature Neuron | brown | original 10, 14
c7 | MV | light blue | original 15

@@

DR: This has changed (slightly). This is the table that I get currently.

Cluster name | Description | Color | Correspondence
-------------|-------------|-------| ----------
c1 | HBC | red | original 1, 5
c2 | GBC | blue | original 2, 3, 11
c3 | mSUS | green | original 4, 7
c4 | mOSN | orange | original 9, 12 
c5 | Immature Neuron | purple | original 10, 14
c6 | mOSN | brown | original 9
c7 | MV | light blue | original 15

@@

DR: I think we should re-write this section. With the current version of the package and parameters, we don't get the contaminant cluster anymore, but most of the contaminant cells are now in c1. However c1 also includes HBC cells, so we should not remove it from the slingshot analysis.

@@

```{r}
pseudoCe <- dataObj[,!primaryClusterNamed(dataObj) %in% c("-1")]
X <- reducedDim(pseudoCe,type="zinbwave")
mds <- cmdscale(dist(X), eig = TRUE, k = 4)
reducedDim(pseudoCe, "MDS") <- mds$points
pseudoCe <- slingshot(pseudoCe, clusterLabels = "combineMany", reducedDim = "MDS", start.clus = "c1")
```
EP: See my email. I don't get this subsetting error on my machine or on TravisCI. I do get on SCF Servers which are running R3.4.1 and the older version of bioconductor. Can you check this on your machine?

@@

DR: Above, I use the syntax to work directly on the ClusterExperiment object, but if we want to be able to plot the resulting lineages, we need the matrix method.

@@

```{r tree, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space, with connecting lines between cluster centers representing the inferred global lineage structure."}
colorCl<-convertClusterLegend(pseudoCe,whichCluster="primary",output="matrixColors")[,1]
lineages <- getLineages(mds$points, clusterLabels = primaryClusterNamed(pseudoCe), start.clus = "c1")
## pairs(lineages, type="lineages", col = colorCl)
```

DR: For some reason, this plot does not work anymore, but the lineages that I get are correct!

@@

```{r fitgam}
t <- colData(pseudoCe)$slingPseudotime_1
t <- slingPseudotime(lineages)[,1]
y <- transformData(pseudoCe)
gam.pval <- apply(y,1,function(z){
  d <- data.frame(z=z, t=t)
  tmp <- gam(z ~ lo(t), data=d)
  p <- summary(tmp)[4][[1]][1,5]
  p
})
```

@@

DR: Kelly, can you please check why I get different lineages and pseudotimes for the slingshot method and for the getLineages/getCurves method? The latter seems the correct one.

```{r heatmapsignificant, fig.cap="DE: Heatmap of the normalized expression measures for the 100 most significantly DE genes for the neuronal lineage, where rows correspond to genes and columns to cells ordered by pseudotime."}
topgenes <- names(sort(gam.pval, decreasing = FALSE))[1:100]

pseudoCe1 <- pseudoCe[,!is.na(t)]
orderSamples(pseudoCe1)<-order(t[!is.na(t)])

plotHeatmap(pseudoCe1[topgenes,], clusterSamplesData = "orderSamplesValue", breaks = .99)
```
